Specialisation \& the Latitude-niche-breadth Hypothesis

### 2015-10-08 manuscript/AC-reply-to-referees_2.tex

\begin{document}

\begin{letter}{\bf Dr. Daniel Costa\\
               Editor\\
               Proceedings of the Royal Society B\\
               6-9 Carlton House Terrace\\
               London, UK\\
               SW1Y 5AG\\
                }

\opening{Dear Dr. Costa:}

Thank you for accepting our manuscript ``\mytitle'', as well as for giving us
the opportunity to address a final round of comments from the reviewers. We
have followed their suggestions and are proud to submit our revision.
Thank you for contributing your time and effort as editor.

@@

\newpage

\setcounter{page}{1}

% -----------------------------------------------------------------------------
% -----------------------------------------------------------------------------
{\Large \bf Reply to Associate Editor}
% ---------------------------------------

  We thank the Associate Editor for his kind words and congratulations, as
  well as for the care and attention shown to our manuscript during the review
  process. We have followed the latest suggestions by the reviewers and are
  happy to acknowledge the Associate Editor and Reviewers' 
  contributions in the manuscript.

{\Large \bf Reply to Reviewer 1}

  We thank the Reviewer for their positive comments and have formally
  acknowledged their contributions in the manuscript. We are pleased that our
  jack-knife analysis has satisfied their concerns about variability between
  studies and appreciate the suggestion that we look into Riede et. al.'s 2011
  results. Although their results for lakes and streams are very interesting
  in the context of our study, our manuscript is already rather lengthy for
  \myjournal. As such, we have included only a very brief reference to Riede
  et. al.'s work. We hope that this will be enough to encourage future
  research exploring the links between specialisation, predator-prey biomass
  ratios, and latitude.

{\Large \bf Reply to Reviewer 2}

  We are pleased that our revised and extended Discussion has met with the Reviewer's approval and appreciate his further minor comments. We have followed all of the Reviewer's suggestions and acknowledge his contributions in the manuscript. With regard to the slight contradiction in the discussion, we have taken care to soften our statements on page 10 to make it clear that there is no latitudinal gradient in our dataset without implying that there is no latitudinal gradient in the true species richness of these ecosystems. We hope that this will remove the apparent contradiction in our discussion.

  \begin{quotation} 
  In either case, the lack of a strong association
  between observed species richness and latitude in any ecosystem type in our dataset
  means that any effect of latitude on other scaling relationships is not being driven by
  the scaling of specialisation with species richness in our food webs.
  \end{quotation}

### 2015-10-09 manuscript/latitudepaper.tex
working on the talk, up to the results more or less

    As the food webs in this dataset are derived from a variety of sources and were compiled over many decades, it
    is likely that they vary in their resolution and in the amount of sampling effort invested in their assembly.
    Many analyses of food-web structure attempt to reduce this variation by using food webs comprised of ``trophic
    species'' --aggregations of species with identical sets of predators and prey-- rather than  
    species \emph{per se}~\cite{Martinez1991,Dunne2004,Vermaat2009,Dunne2013}. As our study
    is concerned directly with the number of species at a particular latitude, however, we did not wish to ignore 
    species with identical sets of interactions. We therefore analysed both original (i.e., without aggregating 
    any species) and trophic-species (i.e., after aggregating species with identical predators and prey) versions
    of the dataset; in each case using the number of species and 
    feeding links in each web to calculate the mean link density (number of links per species), mean generality 
    (number of prey per species), and mean vulnerability (number of predators per species) of the web. 
    The version of the dataset used did not qualitatively change the results, suggesting that
    the scaling relationships between species richness, other food-web properties, and latitude are very 
    similar whether or not species with identical sets of predators and prey are included. For simplicity, 
    here we present only the results for the original (un-aggregated, original species) webs.

@@

    To put our dataset in the context of other research on latitudinal gradients in species richness,
    we first examined simple linear relationships between latitude and each of 
    species richness, links per species, generality, vulnerability, and proportions
    of basal resources, intermediate consumers, and top predators. We fit models of the form

@@

  Link density (mean number of feeding links per species), generality (mean
  number of prey per species), and vulnerability (mean number of predators per
  species) were strongly and positively correlated ($R^2$=0.891 for link
  density and generality, $R^2$\textgreater0.999 for link density and
  vulnerability, and $R^2$=0.890 for generality and vulnerability). Contrary
  to the expected latitudinal gradient, the best-fit version of
  equation~(\ref{Latfull}) did not include a significant
  effect of latitude for any ecosystem type. Further, there were no significant
  relationships between link density, generality, or vulnerability with
  latitude for any ecosystem type.

@@

  The lack of a latitudinal gradient in species richness in this dataset
  contrasts strongly with other
  studies~\cite{Kaufman1995,Macpherson2002,Hillebrand2004,Schemske2009}. As
  numbers of species and links included in a food web vary strongly with
  sampling effort as well as with the underlying diversity of the study area,
  it is possible that the lack of latitudinal trends here is a result of
  researchers tending to expend similar amounts of sampling effort across
  studies. This could result in food webs describing species-rich tropical
  communities omitting more species and links than studies of species-poor
  arctic communities if research groups spend similar person-hours assembling
  webs and can observe similar numbers of species and links per person-hour.
  In addition, it is worth noting that gradients in species richness are
  generally measured for a single taxonomic group at a
  time~\cite{Kaufman1995,Macpherson2002,Hillebrand2004,Schemske2009}. It is
  possible that these taxa are not well-represented in our food webs and that
  the dominant taxa that are represented do not have an underlying latitudinal
  gradient in richness. In either case, the lack of a strong association
  between species richness and latitude in any ecosystem type means that any
  effect of latitude on other scaling relationships is not being driven by an
  underlying latitudinal gradient in species richness. This is fortunate since
  the lack of confounding effects of latitude allows us to more clearly assess
  effects of latitude on scaling with species richness.

@@

  Importantly, while terrestrial communities are also strongly seasonal at
  high latitudes and can receive significant allochthonous
  inputs~\cite{Nakano2001}, terrestrial consumers tend to be
  morphologically specialised for feeding on particular prey~\cite{Liem1990}.
  This means that primarily gape-limited aquatic consumers tend to be more
  generalist across all types of aquatic habitats than terrestrial consumers~\cite{Liem1990,Shurin2006}. The
  key to this explanation of the differences between freshwater and
  marine and estuarine  ecosystems is whether the former experience more
  severe seasonal variation. Although we are not aware of any study explicitly
  comparing seasonal variation in freshwater and saltwater or brackish food
  webs in a similar location, we believe that freshwater ecosystems are indeed
  likely to  experience more severe changes because of their small size. While
  oceans and estuaries certainly vary in terms of water temperature and
  nutrients over the course of a year~\cite{Baird1989}, these changes are likely to be slower
  and milder than in freshwaters because marine and estuarine communities are buffered 
  by being open to the ocean rather than isolated
  in the midst of a terrestrial matrix. Net primary productivity in particular is
  much more variable over the course of a year in non-marine communities~\cite{Field1998},
  suggesting that niche breadths may also be more variable over the course of the year.

@@

  Overall, our results were inconsistent with the latitude-niche breadth
  hypothesis in estuarine, marine, and terrestrial communities but consistent
  with the hypothesis of greater specialisation in the tropics in stream and
  lake food webs. This suggests that different mechanisms may structure food
  webs in different habitat types and that freshwater food webs in particular
  may be strongly affected by seasonal variation. In addition, different
  relationships between latitude and niche breadth in different habitat types
  goes some way towards explaining the equivocal support for the opposing
  hypotheses of narrower niches in the tropics~\cite{Vazquez2004} and broader
  niche space in the tropics~\cite{Davies2007}. Our study indicates that both 
  have merit but would appear to apply to different systems.

@@

  We thank members of the Romanuk, Stouffer, and Tylianakis labs for their comments on
  the manuscript. We also thank Angus McIntosh for valuable discussion on stream food webs.

>>>>

    As the food webs in this dataset are derived from a variety of sources and were compiled over many decades, it
    is likely that they vary in their resolution and in the amount of sampling effort invested in their assembly.
    Many analyses of food-web structure attempt to reduce this variation by using food webs comprised of ``trophic
    species'' --aggregations of species with identical sets of predators and prey-- rather than  
    species \emph{per se}~\cite{Martinez1991,Dunne2004,Vermaat2009,Dunne2013}. As our study
    is concerned directly with the number of species at a particular latitude, however, we did not wish to ignore 
    species with identical sets of interactions. We therefore analysed both original (i.e., without aggregating 
    any species) and trophic-species (i.e., after aggregating species with identical predators and prey) versions
    of the dataset; in each case using the number of species and 
    feeding links in each web to calculate the mean link density (number of links per species), mean generality 
    (number of prey per species), and mean vulnerability (number of predators per species) of the web. 
    The version of the dataset used did not qualitatively change the results, suggesting that
    the scaling relationships between species richness and specialisation across ecosystem type and latitude are very 
    similar whether or not species with identical sets of predators and prey are included. For simplicity, 
    here we present only the results for the original webs.

@@

    To put our dataset in the context of other research on latitudinal gradients in species richness,
    we first examined simple linear relationships between latitude and each of 
    species richness, links per species, generality, and vulnerability. We fit models of the form

@@

  Link density (mean number of feeding links per species), generality (mean
  number of prey per species), and vulnerability (mean number of predators per
  species) were strongly and positively correlated ($R^2$=0.891 for link
  density and generality, $R^2$\textgreater0.999 for link density and
  vulnerability, and $R^2$=0.890 for generality and vulnerability). Contrary
  to the expected latitudinal gradient, the best-fit version of
  equation~(\ref{Latfull}) did not include a significant
  effect of latitude on species richness for any ecosystem type. Further, there were no significant
  relationships between link density, generality, or vulnerability with
  latitude for any ecosystem type.

@@

  The lack of a latitudinal gradient in species richness in this dataset
  contrasts strongly with other
  studies~\cite{Kaufman1995,Macpherson2002,Hillebrand2004,Schemske2009}. As
  numbers of species and links included in a food web vary strongly with
  sampling effort as well as with the underlying diversity of the study area,
  it is possible that the lack of latitudinal trends here is a result of
  researchers tending to expend similar amounts of sampling effort across
  studies. This could result in food webs describing species-rich tropical
  communities omitting more species and links than studies of species-poor
  arctic communities if research groups spend similar person-hours assembling
  webs and can observe similar numbers of species and links per person-hour.
  In addition, it is worth noting that gradients in species richness are
  generally measured for a single taxonomic group at a
  time~\cite{Kaufman1995,Macpherson2002,Hillebrand2004,Schemske2009}. It is
  possible that these taxa are not well-represented in our food webs and that
  the dominant taxa that are represented do not have an underlying latitudinal
  gradient in richness. In either case, the lack of a strong association
  between observed species richness and latitude in any ecosystem type in our dataset
  means that any effect of latitude on other scaling relationships is not being driven by
  the scaling of specialisation with species richness in our food webs. This is fortunate since
  the lack of confounding effects of latitude allows us to more clearly assess
  effects of latitude on scaling with species richness.

@@

  Importantly, while terrestrial communities are also strongly seasonal at
  high latitudes and can receive significant allochthonous
  inputs~\cite{Nakano2001}, terrestrial consumers tend to be
  morphologically specialised for feeding on particular prey~\cite{Liem1990}.
  This means that primarily gape-limited aquatic consumers tend to be more
  generalist across all types of aquatic habitats than terrestrial 
  consumers~\cite{Liem1990,Shurin2006}. This trend also held in our data (
  $\mu_{Aquatic}$=5.47, $\mu_{Terrestrial}$=3.82; $p$=0.007 for $\mu_{Aquatic}$\textgreater$\mu_{Terrestrial}$). 
  The key to this argument is therefore whether freshwater ecosystems experience
  more severe seasonal variation than marine and estuarine ecosystems. Although
  we are not aware of any study explicitly comparing seasonal variation in
  fresh and saltwater ecosystems at similar latitudes, we believe that freshwater 
  ecosystems are indeed likely to experience more severe changes because of their small size.
  While oceans and estuaries certainly vary in terms of water temperature and
  nutrients over the course of a year~\cite{Baird1989}, these changes are likely to be slower
  and milder than in freshwaters because marine and estuarine communities are buffered 
  by being open to the ocean rather than isolated in the midst of a terrestrial matrix. 
  Supporting this hypothesis, net primary productivity is much more variable over the course of a year 
  in non-marine communities~\cite{Field1998}. Thus suggests that niche breadths may also be 
  more variable over the course of the year.

@@

  Overall, our results were inconsistent with the latitude-niche breadth
  hypothesis in estuarine, marine, and terrestrial communities but consistent
  with the hypothesis of greater specialisation in the tropics in stream and
  lake food webs. This suggests that different mechanisms may structure food
  webs in different habitat types and that freshwater food webs in particular
  may be strongly affected by seasonal variation. Freshwater food webs also
  appear to have different predator-prey biomass ratios than other ecosystem
  types~\cite{Riede2011}; although it is not clear whether these ratios are related
  to seasonality, this could be a promising avenue for future research. In addition, different
  relationships between latitude and niche breadth in different habitat types
  go some way towards explaining the equivocal support for the opposing
  hypotheses of narrower niches in the tropics~\cite{Vazquez2004} and broader
  niche space in the tropics~\cite{Davies2007}. Our study indicates that both 
  have merit but would appear to apply to different systems.

@@

  We thank members of the Romanuk, Stouffer, and Tylianakis labs for their comments on
  the manuscript. We also thank Angus McIntosh for valuable discussion on stream food webs and Matthias Schleuning, Robby Stoks, and one anonymous reviewer for their insightful questions.

### 2015-10-09 talk/ABC_2015.tex
working on the talk, up to the results more or less

\usepackage[english]{babel}
\usepackage[absolute,overlay]{textpos}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage[T1]{fontenc}
\usepackage{soul}
\usepackage{pdflscape}
\usepackage{multirow}

@@

    \textbf{Alyssa R. Cirtwill}, Daniel B. Stouffer, Tamara N. Romanuk

@@

@@

      \vspace{1cm}
      {\color{white} 23 October 2015 | ABC 2015}
    \end{center}
  \end{frame}

@@

@@

      \textbf{Envionmental stability?} 
      \textbf{Productivity?} 
      Speciation rates?

      \vspace{1cm}
      {\color{white} 23 October 2015 | ABC 2015}

@@

    \footnotesize{Vazquez, D.P. \& Stevens, R.D. 2004. The latitudinal gradient in niche breadth: concepts and evidence. \emph{The American Naturalist} \textbf{164}, E1-19.}
  \end{frame}

@@

    \tiny{Davies \emph{et. al.} 2007. Productivity alters the scale dependence of the diversity-invasibility relationship. \emph{Ecology} \textbf{88}, 1940-1947.}
  \end{frame}

@@

    \tiny{Davies \emph{et. al.} 2007. Productivity alters the scale dependence of the diversity-invasibility relationship. \emph{Ecology} \textbf{88}, 1940-1947.}
  \end{frame}

\section*{Food Webs}
  \begin{frame}{Testing The Hypotheses}

    \begin{center}
      \includegraphics[width=.8\textwidth]{Figures/hypothesis_comparison.eps}

    \begin{columns}
    \column{.5in}
    \column{1.75in}
      - Greater specialisation in the tropics
    \column{.5in}
    \column{1.75in}
      - Similar specialisation at all latitudes
    \column{.5in}
    \end{columns}

@@

  \begin{frame}{Testing The Hypotheses}

    \begin{center}
      \includegraphics[width=.8\textwidth]{Figures/hypothesis_comparison.eps}

    \begin{columns}
    \column{.5in}
    \column{1.75in}
      - More feeding links per species in tropics
    \column{.5in}
    \column{1.75in}
      - Similar feeding links per species at all latitudes
    \column{.5in}
    \end{columns}

@@

  \begin{frame}{Testing The Hypotheses}

@@

    195 food webs \\
    0-78 degrees (absolute latitude) \\
    Mean L/S, Generality, Vulnerability

@@

    195 food webs \\
    0-78 degrees (absolute latitude) \\
    Mean \textbf{L/S}, Generality, Vulnerability

>>>>

\usepackage[english]{babel}
\usepackage[absolute,overlay]{textpos}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage[T1]{fontenc}
\usepackage{soul}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{pgf}
\usepackage{tikz}

@@

    \textbf{Alyssa R. Cirtwill},\\ Daniel B. Stouffer \& Tamara N. Romanuk

@@

      \vspace{.5cm}

      {\color{white}{\Large Latitude-Niche Breadth Hypothesis}}

      \vspace{.5cm}

@@

@@

      \vspace{.5cm}

      {\Large Latitude-Niche Breadth Hypothesis}

      \vspace{.5cm}

      {\color{purple}Envionmental stability? Productivity?}
      Speciation rates?

@@

@@

    \vspace{1.5cm}

    \tiny{Vazquez, D.P. \& Stevens, R.D. 2004. The latitudinal gradient in niche breadth: concepts and evidence. \emph{Am. Nat.}}
  \end{frame}

@@

    \vspace{1.5cm}

    \tiny{Davies \emph{et. al.} 2007. Productivity alters the scale dependence of the diversity-invasibility relationship. \emph{Ecology.}}
  \end{frame}

@@

    \vspace{1.5cm}

    \tiny{Davies \emph{et. al.} 2007. Productivity alters the scale dependence of the diversity-invasibility relationship. \emph{Ecology.}}
  \end{frame}

\section*{Food Webs}
  \begin{frame}{Does niche breadth vary with latitude?}
    \begin{columns}
    \column{.5in}
    \column{1.75in}
      \begin{center}
      \includegraphics[height=1in]{Figures/version1.eps}\\
      \vspace{.5cm}
      Greater specialisation\\in the tropics

      \vspace{.25cm}
      {\color{white}Fewer links per species\\in the tropics}

      \end{center}
    \column{.5in}
    \column{1.75in}
      \begin{center}
      \includegraphics[height=1in]{Figures/version1.eps}\\
      \vspace{.5cm}
      Similar specialisation\\at all latitudes

      \vspace{.25cm}
      {\color{white}Equal links per species\\at all latitudes}

      \end{center}
    \column{.5in}
    \end{columns}

@@

  \begin{frame}{Does niche breadth vary with latitude?}
    \begin{columns}
    \column{.5in}
    \column{1.75in}
      \begin{center}
      \includegraphics[height=1in]{Figures/version1.eps}\\
      \vspace{.5cm}
      Greater specialisation\\in the tropics

      \vspace{.25cm}
      Fewer links per species\\in the tropics

      \end{center}
    \column{.5in}
    \column{1.75in}
      \begin{center}
      \includegraphics[height=1in]{Figures/version1.eps}\\
      \vspace{.5cm}
      Similar specialisation\\at all latitudes

      \vspace{.25cm}
      Equal links per species\\at all latitudes

      \end{center}
    \column{.5in}
    \end{columns}

@@

  \begin{frame}{Does niche breadth vary with latitude?}

@@

    195 food webs \\
    0-78$^\circ$ (absolute latitude) \\
    Mean L/S, Generality, Vulnerability

@@

    195 food webs \\
    0-78$^\circ$ (absolute latitude) \\
    Mean \textbf{L/S}, Generality, Vulnerability

\section*{Scaling}
  \begin{frame}{Does niche breadth vary with latitude?}

    \begin{center}
      \includegraphics*[width=.8\textwidth]{Figures/results/ls_vs_lat_simulated.eps}
    \end{center}

  \begin{frame}{Niche breadth varies with species richness}

    \begin{center}
      \includegraphics*[width=.8\textwidth]{Figures/results/LS_vs_S_fitline_observed.eps}

    \vspace{.5cm}
    {\color{white}... and species richness varies with latitude}
    \end{center}

  \begin{frame}{Niche breadth varies with species richness}

    \begin{center}
      \includegraphics*[width=.8\textwidth]{Figures/results/LS_vs_S_fitline_observed.eps}

      \vspace{.5cm}
    ... and species richness varies with latitude
    \end{center}

  \begin{frame}{Does latitude affect LS vs. S?}
    \begin{center}
      \includegraphics*[width=.8\textwidth]{Figures/results/LS_vs_S_fitline_observed.eps}

      \vspace{.5cm}
    \end{center}
    \hspace{1in}
      $B \approx aS^{\color{purple}b}$

  \begin{frame}{Does latitude affect LS vs. S?}
    \begin{center}
      \includegraphics*[width=.8\textwidth]{Figures/results/LS_vs_S_fitline_observed.eps}

      \vspace{.5cm}
    \end{center}
    \hspace{1in}
      $B \approx aS^{\color{purple}b_0+b_1Latitude}$

  \begin{frame}{Does latitude affect LS vs. S?}
    \begin{center}
      \includegraphics*[width=.8\textwidth]{Figures/results/LS_vs_S_fitline_observed.eps}

      \vspace{.5cm}
    \end{center}
    \hspace{1in}
      $B \approx aS^{\color{purple}b_0+b_1Latitude+b_2Ecotype+b_3Latitude:Ecotype}$

\section*{Results}
  \begin{frame}{Does latitude affect LS vs. S?}
    \begin{center}
      \includegraphics*[width=.8\textwidth]{Figures/results/marginals.eps}

      I want: just LS, maybe 2 columns?
      Maybe focus on Gen instead of LS? Easier to argue?
